---
title: "MAT3375_Assignment-2"
author: "Rahul Atre"
date: "2023-10-05"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


# Exercise 3.1

Consider the National Football League data in Table B.1

a. Fit a multiple linear regression model relating the number of games won to the team's passing yardage ($x_2$), the percentage of rushing plays ($x_7$), and the opponents' yards rushing ($x_8$).

First, we must import and load the data from table.b1 using an R package containing this textbook data sets. We can do this by running the following command on the console

```
install.packages("MPV", repos = "http://cran.us.r-project.org")
library(MPV)
```

```{r include=FALSE}
library(MPV) 
library(ggplot2)
```

Now, let $x_2$ rep. the team's passing yardage, $x_7$ rep. the percentage of rushing plays, $x_8$ rep. the opponents' yards rushing and $y$ rep. the number of games won by a particular team.

```{r}
data(table.b1)
x_2 = table.b1$x2
x_7 = table.b1$x7
x_8 = table.b1$x8
y = table.b1$y
```

Recall that a regression model that involves more than one regressor variable is called a multiple regression model. It is given by $Y_i = \beta_0 + \sum_{k=1}^{p-1} \beta_k X_ik + \epsilon_i$. In this case, we will have 4 parameters since there are 3 regressor variables.

We can calculate the parameters $b_0$ $b_1$ b_2$ $b_3$ using the lm() function to fit the multiple linear model:

```{r}
full_model = lm(y ~ x_2 + x_7 + x_8) 
full_model 
```

From the above function call, we obtain $b_0$ = -1.8084, $b_1$ = 0.0036, $b_2$ = 0.1940, $b_3$ = -0.0048 (rounded to 4 decimal places).

Therefore, the multiple linear regression model will be $y = -1.8084 + 0.0036x_2 + 0.1940x_7 - 0.0048x_8$.

b. Construct the analysis-of-variance table and test for significance of regression. 

Ans: The analysis-of-variance (ANOVA) table can be generated using R:

```{R}
anova(full_model)
```
Looking at the ANOVA table, we can see that the p-value for $x_2$, $x_7$, and $x_8$ are all very close to 0. This indicates that all regressor variables are **statistically significant**.

c. Calculate t statistics for testing the hypothesis $H_0: \beta_2 = 0, H_0: \beta_7 = 0, and H_0: \beta_8 = 0$. What conclusions can you draw out about the roles the variables $x_2$, $x_7$, and $x_8$ play in the model?

To calculate the t-statistics for the hypothesis, we can generate the summary table: 

```{R}
summary(full_model)
```

As stated, the null hypothesis for each is that the beta parameters are equal to 0. If we look at the table, we can see that the t-value for $\beta_2$ is 5.177, $\beta_7$ is 2.198, and $\beta_8$ is -3.771 which are quite small. The p-value for all three parameters are less than 0.05, which is very small. Therefore, we can reject $H_0$ and conclude that the parameters are **significant** and not equal to 0. In larger context, this means that there is a correlation between the number of games won ($y$) and the team’s passing yardage ($x_2$), percentage of rushing plays ($x_7$), opponent's yard rushing ($x_8$). 


d. Calculate $R^2$ and $R^2_{Adj}$ for this model.

From the above summary table we have generated, $R^2$ = 0.7863 = 78.63%, and $R^2_{Adj}$ = 0.7596 = 75.96%.

e. Using the partial F test, determine the contribution of $x_7$ to the model. How is this partial F statistic related to the t test for $\beta_7$ calculated in part c above?

We can perform a partial F test by comparing the full model and reduced model using anova(). The full model is shown above, and the reduced model contains all predictor variables excluding $x_7$. 

```{R}
reduced_model = lm(y ~ x_2 + x_8) #Reduced model that includes everything but x_7
anova(full_model, reduced_model) 
```
From the above function call, we obtain an F-statistic of 4.8324. This value is the square of the t-statistic of $x_7$ in part c.

# Exercise 3.3

Refer to Problem 3.1.

a. Find a 95% CI on $\beta_7$.

Ans: Since we want the CI on the slope $\beta_7$, we can use the formula $b_7 \pm t_{1-\alpha/2, n-2} * s(b_7)$. We know that $n=28$, $b_7 = 0.193960$, $\alpha = 0.95$. We can generate the CI using R:

```{R}
confint(full_model, level = 0.95)
```

Therefore, from the above function call, we are 95% confident that $\beta_7$ is in the interval [0.0119, 0.3761].

b. Find a 95% CI on the mean number of games won by a team when $x_2 = 2300, x_7 = 56.0, x_8 = 2100$.

Ans: In order to find this, we need to set the predictor variables to the given values above to get a 95% confidence interval, and use the point estimate $\hat{Y} = b_0 + b_1x_2 + \beta_2x_7 + \beta_3x_8$. So, the CI is $\hat{Y} \pm t_{1-\alpha/2, n-2} * s(\hat{Y})$. Generating the CI using R:

```{R}
new.dat = data.frame(x_2 = 2300, x_7 = 56.0, x_8 = 2100)
predict(full_model, newdata = new.dat, interval = "confidence")
```

Therefore, a 95% CI on the mean number of games won by team given the above constraints is [6.4362, 7.9966].


# Exercise 3.4

Reconsider the National Football League data from Problem 3.1. Fit a model to these data using only $x_7$ and $x_8$ as the regressors.

- We can create a reduced model that will only contain $x_7$ and $x_8$ as the regressors by applying the same lm() function:

```{r}
reduced_model = lm(y ~ x_7 + x_8) #Excluding x_2
reduced_model 
```

The reduced model is $\hat{Y} = y = 17.9443 + 0.0484x_7 - 0.0065x_8$


a. Test for significance of regression.


The analysis-of-variance (ANOVA) will generate the following output for significance of regression:

```{R}
summary(reduced_model)
```
From the above function call, we can see that the F-statistic for the reduced model is 15.13 which is smaller than 30, and a p-value which is very close to 0. This means that a model with only $x_7$ and $x_8$ is **statistically significant**.


b. Calculate $R^2$ and $R^2_{Adj}$. How to these quantities compare to the values computed for the model in Problem 3.1, which included an additional regressor ($x_2$)?

From the above summary table we have generated, $R^2$ = 0.5477 = 54.77%, and $R^2_{Adj}$ = 0.5115  = 51.15%. These quantities are **close to 20% less** than the model in Problem 3.1, which included $x_2$.


c. Calculate a 95% CI on $\beta_7$. Also find a 95% CI on the mean number of games won by a team when $x_7 = 56.0$ and $x_8 = 2100$. Compare the lengths of these CIs to the lengths of the corresponding CIs from Problem 3.3.

```{R}
confint(reduced_model, level = 0.95)
```

From the above function call, we are 95% confident that $\beta_7$ is in the interval [-0.1972, 0.2939].


```{R}
new.dat = data.frame(x_7 = 56.0, x_8 = 2100)
predict(reduced_model, newdata = new.dat, interval = "confidence")
```
A 95% CI on the mean number of games won by team with the reduced model constraints is [5.8286, 8.0238].

Both the confidence intervals have **higher values** in comparison to the full model (i.e. values that are shifted to a greater value), although from a high-level overview, they seem to have a similar range between the confidence intervals. 

d. What conclusions can you draw from this problem about the consequences of omitting an important regressor from the model?

By removing an important regressor from the model, there is a significant drop in the total variability ($R^2$) explained by the model. In addition, the confidence interval shifted slightly to the right, and the standard errors of coefficients also changed. 


# Exercise 4.1

Consider the simple regression model fit to the National Football League team performance data in Problem 2.1.

a. Construct a normal probability plot of the residuals. Does there seem to be any problem with the normality assumption?

We first need to obtain the residuals of the dataset (same as previous questions):

```{R}
full_model2 = lm(y ~ x_8)
residuals <- resid(full_model2) #Obtain residuals from the full_model
qqnorm(residuals, main = "Normal Probability Plot of Residuals")
qqline(residuals) #Add line of best fit
grid()
```
Above is the graph for the normal probability plot of residuals. Since most of the residual points follow the line of best fit, there appears to be no problem with the normality assumption.

b. Construct and interpret a plot of the residuals versus the predicted response. 

```{R}
ggplot(full_model2, aes(.fitted, .resid)) + geom_point() + geom_hline(yintercept = 0) + labs(title = "Residuals vs. Predicted Response") + xlab("Fitted Value") + ylab("Residuals")
```
Above is the plot of residuals versus the predicted response. Most of the points appear to be close to the horizontal line where residual = 0. Since there is no clear pattern among the points, it implies that the residuals are random and not correlated to the predicted response. 

Thus, we can conclude that the model's predictions are mostly accurate, with some variance that is reasonable.  

c. Plot the residuals versus the team passing yardage $x_2$. Does this plot indicate that the model will be improved by adding $x_2$ to the model?


```{R}
ggplot(full_model2, aes(x_2, .resid)) + geom_point() + geom_hline(yintercept = 0) + labs(title = "Residuals vs. Predicted Response") + xlab("x2") + ylab("Residuals")
```

For the plot with only $x_2$, there is no clear pattern among the points and shows a higher concentration of points around the center of the graph. Thus, we can conclude that when the model only contains $x_2$ as its predictor variable, the model is more accurate. 


```{R}
myplot = mfrow = c(4, 2)
plot(full_model)
```