---
title: "MAT3375_Assignment-1"
author: "Rahul Atre"
date: "2023-09-15"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Exercise 2.1

Table B.1 gives data concerning the performance of the 26 National Football League teams in 1976. It is suspected that the number of yards gained rushing by opponents ($x_8$) has an effect on the number of games won by a team (y).

a.  Fit a simple linear regression model relating games won y to yards gained rushing by opponents $x_8$.

Ans: For this example, we will use the method of least squares to obtain a fitted regression model.

First, we must import and load the data from table.b1 using an R package containing this textbook data sets. We can do this by running the following command on the console

```
install.packages("MPV", repos = "http://cran.us.r-project.org")
library(MPV)
```

```{r include=FALSE}
library(MPV) 
```

Now, let x_yards, $x_8$, rep. the \# of yards gained rushing by opponents, and y_won, $y$ rep. the number of games won by a particular team

```{r}
data(table.b1)
x_yards = table.b1$x8
y_won = table.b1$y
```

The desired simple regression model: $Y_i = \beta_0 + \beta_1 X_i + \epsilon_i$ According to methods of least squares, $b_0 = \overline{Y} - b_1\overline{X}$ and $b_1 = \frac{\sum (X_i - \overline{X})Y_i}{\sum (X_i - \overline{X})^2}$

We can calculate the parameters $b_0$ and $b_1$ using the lm() function to fit the linear model:

```{r}
lin_model = lm(y_won ~ x_yards) 
lin_model 
```

From the above function call, we obtain $b_0$ = 21.788 and $b_1$ = -0.007 (rounded to 3 decimal places).

Therefore, a simple linear model relating games won to yards gained rushing by opponents is $\hat{Y} = 21.788 - 0.007X$.

b.  Construct the analysis-of-variance table and test for significance of regression.

Ans: The analysis-of-variance (ANOVA) table can be generated using R:

```{R}
anova(lin_model)
```

As for testing the significance of regression, the anova table indicates how to test the null hypothesis:

$H_0 : \beta_1 = 0$

$H_1 : \beta_1 \neq 0$

As we know, the null hypothesis is that the slope of the line is equal to 0. However, if we take a look at the F and p-value, we can see that the F is 31.103 which is quite large, and the p-value is less than 0.001, which is very small. Therefore, we can reject $H_0$ and conclude that the slope is not equal to 0. In larger context, this means that there is a correlation between yards gained rushing by opponents and the number of games won by a team.

c.  Find a 95% CI on the slope.

Ans: Since we want the CI on the slope $\beta_1$, we can use the formula $b_1 \pm t_{1-\alpha/2, n-2} * s(b_1)$. We know that $n=28$, $b_1 = -0.007$, $\alpha = 0.95$. We can generate the CI using R:

```{R}
confint(lin_model, level = 0.95)
```

Therefore, from the above function call, we are 95% confident that the slope is in the interval [-0.0096, -0.0044].

d.  What percent of the total variability in y is explained by this model?

Ans: The coefficient of determination $R^2$ can be used to find the percentage of total variability in y from the formula $$\frac{\sum (\hat{Y} - \overline{Y})^2}{\sum (Y_i - \overline{Y})^2}$$. Using R to generate this:

```{R}
summary(lin_model)
```

From the above function call, $R^2$ = 0.5447 = 54.47%. Therefore, about 54.47% of the total variability in y is explained by this model.

e.  Find a 95% CI on the mean number of games won if opponents yards rushing is limited to 2000 yards.

Ans: In order to find this, we need to set $X = 2000$ to get a 95% confidence interval, and use the point estimate $\hat{Y} = b_0 + b_1X$. So, the CI is $\hat{Y} \pm t_{1-\alpha/2, n-2} * s(\hat{Y})$. Generating the CI using R:

```{R}
new.dat = data.frame(x_yards = 2000)
predict(lin_model, newdata = new.dat, interval = "confidence")
```

Therefore, a 95% CI on the mean number of games won if opponents yards rushing is limited to 2000 yards is [6.7658, 8.7103].

# Exercise 2.10

The weight and systolic blood pressure of 26 randomly selected males in the age group 25-30 are shown below. Assume that weight and blood pressure (BP) are jointly normally distributed.

a.  Find a regression line relating systolic blood pressure to weight.

Ans: For this example, we will use the method of least squares to obtain the estimates for $\beta_0$ and $\beta_1$.

Let x_weight rep. the weight of a male. Let y_systolicBP represent the systolic blood pressure of a male.

We must first load the data from the table. Then, we can use the lm() function to find the regression line in R:

```{R}
data(p2.10)
x_weight = p2.10$weight
y_systolicBP = p2.10$sysbp

lin_model = lm(y_systolicBP ~ x_weight)
lin_model
```

From the above function call, we obtain $b_0$ = 69.1044 and $b_1$ = 0.4194.

Therefore, a regression line relating systolic blood pressure to weight is $\hat{Y} = 69.1044 + 0.4194X$.

b.  Estimate the correlation coefficient.

Ans: The sample correlation coefficient is defined as the following:

$$\frac{\sum(x_i - \overline{x})(y_i - \overline{y})}{\sqrt{\sum(x_i - \overline{x})^2\sum(y_i - \overline{y})^2}}$$. Using R to generate this:

```{R}
cor(x_weight, y_systolicBP)
```

From the above function call, we obtain $r = 0.7735$. Therefore, the estimated correlation coefficient is 0.7735.

c.  Test the hypothesis that $\rho = 0$.

Ans: We have the following hypothesis for $\rho$:

$H_0 : \rho = 0$

$H_1 : \rho \neq 0$

The following test statistic is given for $\rho$ under $H_0$:

$t = \frac{r\sqrt{n-2}}{\sqrt{1-r^2}} \sim Student_{n-2}$. Using R to generate this:

```{R}
cor.test(x_weight, y_systolicBP)
```

We can see that the t-statistic is 5.9786 which is quite small, and the p-value is less than 0.001, which is very small. Therefore, we can reject $H_0$ and conclude that the correlation coefficient is not equal to 0. In larger context, this means that there is evidence to suggest a correlation between the systolic blood pressure and weight.

d.  Test the hypothesis that $\rho = 0.6$.

Ans: We have the following hypothesis for $\rho$:

$H_0 : \rho_0 = 0.6$

$H_1 : \rho_0 \neq 0.6$

The test-statistic for $H_0 : \rho = \rho_0$ can be computed as shown below. We know that $r = 0.7735$, $\rho_0 = 0.6$, $n = 26$:

$Z = (arctanh(r) - arctanh(\rho_0))(n-3)^{1/2}$

$Z = (arctanh(0.7735) - arctanh(0.6))(26-3)^{1/2}$ $Z = (1.02898 - 0.693147)(23)^{1/2}$ $Z = 1.61061$

If we let $\alpha = 0.5$, then the rejection for $H_0$ is if $|Z_0| > Z_{\alpha/2} = 1.96$. Since that is not true, we do not reject H_0, as there is not enough evidence to suggest that $\rho \neq 0.6$, $H_1$, is true.

e.  Find a 95% CI for $\rho$.

Letting $\alpha = 0.5$, the formula for a 95% CI of $\rho$ is given as follows:

$[tanh(arctanh(r) - \frac{Z_{\alpha/2}}{\sqrt{n-3}}), tanh(arctanh(r) + \frac{Z_{\alpha/2}}{\sqrt{n-3}})]$. Using R to generate this:

```{R}
cor.test(x_weight, y_systolicBP)
```

From the above function call, we get the 95% CI interval for $\rho$ to be [0.5513, 0.8932].


# Exercise 2.22

Consider the methanol oxidation data in Table B.20. The chemist believes that the ratio of inlet oxygen to the inlet methanol controls the conversion process. Perform a through analysis of these data. Do the data support the chemist's belief?

Ans: First, let's import the data from table.b20 in the textbook data set.

Let x_OxyMethRatio, $x_5$ rep. the \# of ratio of inlet oxygen to inlet methanol, and y_convProcess, $y$ rep. the percent conversion.

```{r}
data(table.b20)
x_OxyMethRatio = table.b20$x5
y_convProcess = table.b20$y
```

Using the lm() and summary function to obtain a summary of the data:

```{R}
lin_model = lm(y_convProcess ~ x_OxyMethRatio)
summary(lin_model)
```

From the above function call, we obtain $\hat{Y} = 21.25 + 7.80x_5$. 

Testing the null hypothesis, $H_0: \beta_1 = 0$ against $H_1: \beta_1 \neq 0$, we obtain F to be 0.2161 and p-value as 0.6483. Also, $R^2 = 0.0133$, meaning that about 1.3% of the total variability in Y is explained by the model, which is not much. Since the F and p-value is small, we fail to reject $H_0$.

Therefore, this implies that there is not enough evidence to suggest the ratio of inlet oxygen to inlet methanol controls the conversion process. The data does not support the chemist's beliefs.
